---
title: "Playlist #437:Songs My Algorithm Thinks You'll Like"
---

Remember to replace all of the instructions with your text!

## Summary

This section should include:

-   Your main question or objective
-   The main insight(s) from your project
-   A figure that supplements that insight, along with accompanying text describing the figure. This figure can be a cleaned-up version of a visualization you produced in EDA or a visualization using your final model.

### Main Question and Objective

This project investigates what audio characteristics make a Spotify song popular and whether those songs can be meaningfully groups and classifies by genre based on those characteristics terms. Using a subset of the Spotify Song Dataset refined to Pop, R&B, and EDM genres, the three most streamed genres and my personal favorites, the project explores how measurable audio features such as loudness, energy, dancebility, tempo, and acousticness correlates to a song's popularity score and track genre.

### Main Insights

Across the exploratory, unsupervised, and supervised analyses, three consistent findings emerged that together answer the project's central question. Audio features such as energy, danceability, and loudness all showed a modest positive relationship with track popularity that plateaued at extreme values, suggesting that sufficiently high production intensity gives songs a streaming advantage within Pop, R&B, and EDM, but that no single audio feature is a strong standalone predictor of success.These same audio dimensions proved partially useful for predicting genre in supervised classification, with the decision tree achieving approximately 62% cross-validated accuracy well above the 33% random baseline, though Pop remained the hardest genre to classify correctly because it borrows stylistic elements from both EDM and R&B and occupies the most central and overlapping position in audio feature space. Taken together, these findings suggest that while Spotify's audio features capture meaningful structure in music, popularity and genre identity are both complex outcomes that audio characteristics alone can only partially explain.


## Motivation and Context
Music streaming has fundamentally changed how songs become popular. On platforms like Spotify, which hosts over 100 million tracks and serves more than 600 million users, a song's visibility is shaped not just by marketing or radio play but by algorithmic recommendation systems that match listeners to music based on audio characteristics. Understanding what makes a song popular — and what audio features define genre boundaries — has practical implications for artists, producers, and playlist curators alike.
This project uses the Spotify Song Dataset, a publicly available collection of track metadata and audio features extracted through the Spotify Web API, to investigate two related questions: what audio characteristics are associated with higher popularity scores, and whether those same characteristics can be used to classify songs by genre without any prior genre information. By focusing on Pop, R&B, and EDM — three genres that together dominate modern streaming — the analysis explores both the predictive power of Spotify's audio features and the degree to which genre identity is encoded in measurable sound characteristics.

### Insigh Figure

```{r}
#| label: do this first
#| echo: false
#| message: false

here::i_am("math-437-project-template.qmd")
```

This section describes what you are investigating in your project and why you are investigating it. You should provide enough contextual background and information that someone with a limited background can understand the broad outlines of the topic being investigated.

## Packages Used In This Analysis

This section lists the packages used in the analysis. You should have a single chunk that loads the packages followed by a description of why you are using each package. Please use individual package names rather than tidyverse and tidymodels, to make it clear which specific packages are being used.

```{r}
#| label: load packages
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(tidyclust)
library(tidyr)
library(GGally)
library(discrim)
library(rpart)
library(mvtnorm)
```

You should follow the template table below. Each row of the table contains the package name in brackets, a link to the package website, and (on the right side) a description of what you used the package for. If you can't find a link to the package website, or you're having trouble getting the link to work, don't worry about including it.

| Package | Use |
|--------------------------------|----------------------------------------|
| [here](https://github.com/jennybc/here_here) | to easily load and save data |
| [readr](https://readr.tidyverse.org/) | to import the CSV file data |
| [dplyr](https://dplyr.tidyverse.org/) | to massage and summarize data |
| [rsample](https://rsample.tidymodels.org/) | to split data into training and test sets |
| [ggplot2](https://ggplot2.tidyverse.org/) | to create nice-looking and informative graphs |
| [tidyverse](https://www.tidyverse.org) | to provide a collection of core data science packages for data import, wrangling, and visualization |
| [tidymodels](https://tidymodels.tidymodels.org) | to provide a unified framework for modeling and machine leanring workflows |
| [tidyclust](https://tidyclust.tidymodels.org/) | to perform clustering within the tidymodels framework |
| [tidyr](https://tidyr.tidyverse.org) | to reshape and tidy data into a consistent format |
| [GGally](https://ggobi.github.io/ggally/) | to create extended visualizations such as pair plots |
| [discrim](https://discrim.tidymodels.org/) | to access classification models like discriminant analysis |
| [rpart](https://cran.r-project.org/package=rpart) | to build recursive partitioning trees for regression and classification |
| [mvtnorm](https://cran.r-project.org/package=mvtnorm) | to compute multivariate normal and t-distribution functions |

## Data Description

This section should describe the data that you are using, the source of the data (including a link if appropriate), and how the data was originally collected. You should be able to just clean up the Data Description section from Project 2.

At the end of the section you should import the data, unless it's already in one of the packages you loaded.

The dataset used for this project is the **Spotify Song Data**, which contains information about available songs on the music streaming platform Spotify. For this dataset, each observation represents one individual song that appears in a Spotify playlist. The original dataset contains 32,833 observations and 23 variables. These variables include identifying information such as track name, artist name, album name, and playlist genre, along with numerical audio features measured by Spotify's internal algorithms and documented through the Spotify Web API. These audio features are defined by Spotify as follows: ***danceability*** measures how suitable a track is for dancing based on tempo, rhythm stability, and beat strength on a scale of 0 to 1; ***energy*** represents the perceived intensity and activity of a track, where higher values correspond to faster, louder, and more dynamic songs; ***valence*** describes the musical positivity of a track, with higher values sounding more happy and euphoric; ***tempo*** measures the estimated beats per minute of the track; ***acousticness*** measures the likelihood that a track was produced without electronic amplification on a scale of 0 to 1; ***speechiness*** detects the presence of spoken words in a track; ***loudness*** measures the overall average volume of a track in decibels, and ***playlist_genre*** that categorizes each song to 1 of 7 genres, or in the case of the subsetted data, 1 of 3 genres. The response variable of interest is ***track_popularity***, which Spotify defines as a score between 0 and 100 calculated primarily by total number of plays and how recent those plays are, where 100 represents the most popular songs on the platform.

The dataset was compiled from Spotify using the `spotifyr` package and the Spotify Web API, which provides public access to song metadata, playlist information, and audio characteristics. The data was collected by queueing playlists and extracting track level information, making this dataset possible for predictive modeling as it contains streaming data with documentation for each variable. The data was imported into R using the `read_csv()` function.

```{r}
#| label: import data
#| warning: false
spotify_songs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-01-21/spotify_songs.csv')
```

## Data Wrangling (Optional Section)

Since the full dataset includes many genres, the data was subset to only include songs from the Pop, R&B, and EDM genres, reducing the dataset down to 16,981 observations. These genres were selected because they are more represented in modern streaming music and provide enough observations for meaningful analysis. These genres are also my personal top genres to listen to on Spotify. This subset allows for easier comparison of song characteristics while reducing noise from less common genres.

After importing, the subsetted dataset was divided into a training set of 70% and a test set of 30% using random sampling. This split is commonly used in statistical learning because it provides enough observations to train accurate models while reserving a substantial portion of unseen data for evaluating predictive performance. The training set is used to build models, while the test set gives an unbiased measure of how well the model generalizes to new songs.

Now, using this dataset, we can answer an overall question: What characteristics make a Spotify song popular, and can songs be grouped and predicted based on audio filter?

```{r}
#| label: subset spoitify songs by genre
spotify <- spotify_songs[spotify_songs$playlist_genre %in% c("pop", "edm", "r&b"), ]
```

```{r}
#| label: split into training and testing set

set.seed(127, kind = "Mersenne-Twister")
spotify_split <- initial_split(spotify,
                               prop = 0.70
                               )
spotify_train <- training(spotify_split)
spotify_test <- testing(spotify_split)

```

## Exploratory Data Analysis

This section should include the code, figures, and accompanying text for the exploratory data analysis that is relevant to your objective. You should be able to just clean up the EDA section from Project 2.


### Track Popularity
Track popularity is the main response variable in this project making it important to first understand how popularity score are distributed across songs in the training set. By examining the distribution, it helps determine whether most songs are similarly popular or if a small number of songs are scaled on a relatively higher popularity score.

```{r}
#| label: popularity distribution
ggplot(data = spotify_train,
       mapping = aes(x = track_popularity)
       ) + 
  geom_histogram(binwidth = 5,
                 fill = "skyblue",
                     color = "black"
                 ) + 
  labs(title = "Distribution of Song Popularity",
       x = "Track Popularity",
       y = "Count")
```

The histogram supports these findings by showing that popularity is spread across the range, with many songs concentrated in the low to middle and middle ranges rather than only at the highest values. There are fewer songs near the maximum popularity levels, suggesting that extremely popular songs are less common than moderately popular ones. This pattern is realistic because only a limited number of songs become major hits, while many others achieve moderate visibility.


### Danceability

Danceability is next explored as it is one of Spotify's most recognizable features and may help to explain why some songs are more popular than others. According to Spotify, danceability measures how suitable a track is for dancing based on characteristics such as tempo, rhythm stability, beat strength, and overall regularity. Since the dataset focuses on Pop, Rap, and EDM songs, danceability is especially relevant because these genres often emphasize rhythm and movement.

```{r}
#| label: danceability distribution 
ggplot(data = spotify_train,
       mapping = aes(x = danceability)
       ) + 
  geom_histogram(binwidth = 0.05,
                 fill = "skyblue",
                     color = "black"
                 ) + 
  labs(title = "Distribution of Song Danceability",
       x = "Danceability",
       y = "Count")
```

The histogram showcases these findings by showing that most songs are concentrated between roughly 0.55 and 0.80, with relatively few songs at the very low end of the scale. This pattern is expected because Pop, R&B, and EDM are genres commonly associated with strong beats and energetic production. Songs with extremely low danceability are uncommon in this subset.


### Popularity vs. Danceability

The relationship between track popularity and dancebility is compared to determine whether song with lower danceability tend to perform better on Spotify streaming. Since danceability reflects rhythm strength, beat consistency, and movement potential, it is reasonable to investigate whether more danceable songs attract more listeners and streams, especially within Pop, R&B, and EDM genres.

```{r}
#| label: scatterplot of danceability of popularity
ggplot(data = spotify_train,
       mapping = aes(x = danceability,
                     y = track_popularity)
       ) +
  geom_point(alpha = 0.2,
             color = "skyblue4") +
  geom_smooth(method = "loess",
              se = FALSE) +
  labs(title = "Popularity vs. Danceability",
       x = "Danceability",
       y = "Track Popularity")
```


## Modeling

The section should include the code, output, and accompanying descriptions for *one* supervised or unsupervised learning model, as appropriate to your question or objective. The descriptions should include some conceptual or mathematical background explaining how the model works.

You should be able to just clean up one of the modeling sections from Project 2.

## Insights/Conclusion

Pick either "Insights" or "Conclusion" as the section title depending on what makes more sense.

Describe the primary findings from your EDA and modeling sections as appropriate to the main question/objective. You can reproduce the figure from your summary if that helps.

### Limitations and Future Work

This section should include further exploration of the limitations of your modeling approach. Some avenues of exploration include:

-   Who/what is being poorly predicted by your model? Why?
-   How might someone improve the data collection or analysis in the future?
-   Are the assumptions behind your analytical methods justified?
-   Are there ethical concerns about how the data was collected or how the results of the analysis could be used in the real world?
